---
title: "Hartshorne Tenenbaum Pinker (2018) Spotligth"
author: "Mike Frank"
date: "May 5th, 2018"
output: html_document
---

This document contains some explorations of HTP's 2018 paper on critical periods, including the reproducibility code for a figure in my TiCS spotlight paper. 

Note that it depends on data from the HTP paper OSF repository and also on running a fairly time-consuming MIRT model. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggrepel)
library(mirtCAT)
library(viridis)
library(ggthemes)
```

# Initial data evaluation

```{r}
d <- read_csv("data.csv")
```


Number of people at ceiling. 

```{r}
mean(d$correct >= .96)
```


```{r}
qplot(d$age)
```

```{r}
qplot(d$Eng_start)
```

```{r}
qplot(d$Eng_country_yrs)
```

# Basic reproduction of plots

break down by monolingual, immersion, non-immersion

```{r}
ms <- d %>%
  mutate(start_binned = cut(Eng_start, c(0,seq(1,50,5), 75), 
                            include.lowest = TRUE)) %>%
  group_by(age, start_binned) %>%
  summarise(correct = mean(correct), 
            n = n())

ggplot(ms,  
       aes(x = age, y = correct, col = start_binned)) +
  geom_line() + 
  geom_point(aes(size = n), alpha = .2) + 
  # geom_smooth(aes(group = start_binned), se=FALSE, span = 1) + 
  theme_few() 
```

Just monolinguals

```{r}
ms <- d %>%
  mutate(status = ifelse(natlangs == "English" & primelangs == "English" &
         Eng_start == 0, "monolingual","non")) %>%
  group_by(age, gender, education, Eng_little) %>%
  summarise(correct = mean(correct), 
            n = n()) %>%
  filter(n > 30)


ggplot(filter(ms, gender %in% c("female","male")),  
       aes(x = age, y = correct, col = education)) +
  # geom_line() + 
  geom_point(aes(size = log10(n)), alpha = .2) + 
  facet_grid(Eng_little~gender) +
  geom_smooth(aes(group = education), se=FALSE, span = 1) +
  theme_few() +
  scale_color_discrete(name = "Initial Exposure Age") +
  xlab("Age") + 
  ylab("Mean Correct") +
  theme(legend.position = "bottom")
```

# Psychometric modeling

Fit 4 parameter IRT model to full dataset via mirtCAT (cached, takes many minutes). 

```{r, eval = FALSE}
library(mirtCAT)

d_mat <- d %>%
  select(starts_with("q")) %>% 
  data.frame %>%
  data.matrix

d_mat <- d_mat[complete.cases(d_mat),]

mod_4pl <- mirt(d_mat, 1, itemtype='4PL', verbose=TRUE)

coefs_4pl <- as_data_frame(coef(mod_4pl, simplify = TRUE)$items)
fscores_4pl <- fscores(mod_4pl, method = "MAP")[,1]

save(file = "mod_4pl.Rds", "mod_4pl","fscores_4pl", "coefs_4pl")

```

Load cache. 

```{r}
load("mod_4pl.Rds")
d$fscore <- fscores_4pl
```

Examine item performance.

```{r}
coefs_4pl$name <- d %>% select(starts_with("q")) %>% names()

ggplot(coefs_4pl,  
       aes(x = a1, y = d)) + 
  geom_point() + 
  geom_label_repel(data = filter(coefs_4pl, a1 > 10 | d > 10), 
                                 aes(label = name)) + 
  xlab("Discrimination") + 
  ylab("Difficulty")
```

```{r}
ggplot(coefs_4pl,  
       aes(x = g, y = u)) + 
  geom_point() + 
  geom_label_repel(data = filter(coefs_4pl, u < .92), 
                                 aes(label = name)) + 
  xlab("Lower bound (guessing)") + 
  ylab("Upper bound (ceiling)") + 
  xlim(0,1) + 
  ylim(0,1)
```

Look at item characteristic curves for some of the odd items.

```{r}
itemplot(mod_4pl, which(coefs_4pl$name %in% "q2"))
itemplot(mod_4pl, which(coefs_4pl$name %in% "q34_4"))
itemplot(mod_4pl, which(coefs_4pl$name %in% "q33_5"))
itemplot(mod_4pl, which(coefs_4pl$name %in% "q29_3"))
itemplot(mod_4pl, which(coefs_4pl$name %in% "q29_1"))
itemplot(mod_4pl, which(coefs_4pl$name %in% "q1"))
```

Look at correct answer distributions

```{r}
d <- d %>%
  mutate(education = factor(education,
                            levels = c("Graduate Degree",
                                       "Some Graduate School",
                                       "Undergraduate Degree (3-5 years higher ed)",
                                       "Some Undergrad (higher ed)",
                                       "High School Degree (12-13 years)",
                                       "Didn't Finish High School (less than 13 years ed)"),
                            labels = c("Graduate","Some Graduate","College", 
                                       "Some College",
                                       "High School","< High School")))

d_hist <- d %>%
  mutate(Eng_little = ifelse(!is.na(Eng_little), Eng_little, "other"), 
         language_status = fct_relevel(
           fct_recode(Eng_little, 
                      Bilingual = "bileng",
                      Monolingual = "monoeng", 
                      `High Immersion` = "lot",
                      `Low Immersion` = "little", 
                      `Moderate Immersion` = "other"),
           "Monolingual","Bilingual", 
           "High Immersion", 
           "Moderate Immersion","Low Immersion"))
                
ggplot(d_hist, 
       aes(x = correct, fill = language_status)) + 
  geom_histogram() + 
  theme_few() + 
  scale_fill_ptol(name = "Language Status") + 
  scale_y_continuous(name = "Number of participants", 
                     breaks = c(1e5,2e5,3e5,4e5)) +
  xlab("Proportion correct") + 
  theme(legend.position = "bottom") + 
  guides(fill=guide_legend(ncol=3))
```

and latent ability. 

```{r}
ggplot(d_hist, 
       aes(x = fscore, fill = language_status)) + 
  geom_histogram(binwidth = .25) + 
  theme_few() + 
  scale_fill_ptol(name = "Language Status") + 
  scale_y_continuous(name = "Number of participants", 
                     breaks = c(1e5,2e5,3e5,4e5)) +
  xlab("Latent ability") + 
  xlim(-3.5, .8) + 
  theme(legend.position = "bottom") + 
  guides(fill=guide_legend(ncol=3))
```


Now facet by language status. First show proportion correct. 

```{r}
ggplot(d_hist,
       aes(x = correct, fill = language_status)) + 
  geom_histogram() + 
  facet_wrap(~ language_status, scales = "free_y") + 
  theme_few() + 
  scale_fill_ptol(name = "Language Status") + 
  scale_y_continuous(name = "Number of participants") +
  xlab("Proportion correct") + 
  theme(legend.position = "bottom") + 
  guides(fill = guide_legend(ncol = 3))
```

Then F score. 

```{r}
ggplot(d_hist,
       aes(x = fscore, fill = language_status)) + 
  geom_histogram() + 
  facet_wrap(~language_status, scales = "free_y") + 
  theme_few() + 
  scale_fill_ptol(name = "Language Status") + 
  scale_y_continuous(name = "Number of participants") +
  xlab("Latent ability") + 
  theme(legend.position = "bottom") + 
  guides(fill=guide_legend(ncol = 3))
```

By education.

```{r}
ggplot(filter(d_hist, !is.na(education)),
       aes(x = fscore, fill = education)) + 
  geom_histogram(binwidth = .25) + 
  facet_wrap(~language_status, scales = "free_y") + 
  theme_few() + 
  scale_fill_viridis(name = "Language Status", discrete = TRUE) + 
  scale_y_continuous(name = "Number of participants") +
  xlim(-3.5,.8) + 
  xlab("Latent ability") + 
  theme(legend.position = "bottom") + 
  guides(fill=guide_legend(ncol = 3)) 
```



```{r}
d_hist_long <- d_hist %>% 
  select(fscore, correct, language_status) %>%
  gather(measure, value, fscore, correct) %>%
  mutate(measure = factor(measure, 
                        levels = c("correct","fscore"), 
                        labels = c("Raw proportion correct", "Latent ability (IRT)")))

ggplot(d_hist_long,
       aes(x = value, fill = language_status)) + 
  geom_histogram() + 
  facet_grid(measure~language_status, scales = "free_x") + 
  theme_few() + 
  scale_fill_ptol(name = "Language Status") + 
  scale_y_continuous(name = "Number of participants") +
  xlab("Measure") + 
  theme(legend.position = "bottom") + 
  guides(fill=guide_legend(ncol=3))
```



Plot latent abilities in the same way we plot the data generally. 

```{r}
ms <- d %>%
  mutate(start_binned = cut(Eng_start, c(0, 1, 5, 10, 20, 100), 
                            include.lowest = TRUE), 
         immersion_pct = ifelse(!is.na(Eng_country_yrs), 
                                Eng_country_yrs / (age - Eng_start), 1),
         immersion_binned = cut(immersion_pct, c(0,.1, .9, 1.01), 
                            include.lowest = TRUE)) %>%
  group_by(age, start_binned, immersion_binned) %>%
  summarise(theta = mean(fscore), 
            n = n()) %>%
  filter(n > 30)

ggplot(ms,  
       aes(x = age, y = theta, col = start_binned)) +
  # geom_line() + 
  geom_point(aes(size = log10(n)), alpha = .2) + 
  facet_wrap(~immersion_binned) + 
  geom_smooth(aes(group = start_binned), se=FALSE, span = .75) +
  theme_few() +
  scale_color_discrete(name = "Initial Exposure Age") +
  xlab("Age") + 
  ylab("Latent Ability (IRT theta)") +
  theme(legend.position = "bottom")
```

```{r}
ms <- d %>%
  mutate(start_binned = cut(Eng_start, c(0, 1, 5, 10, 20, 100), 
                            include.lowest = TRUE), 
         Eng_little = ifelse(!is.na(Eng_little), Eng_little, "other"), 
         language_status = fct_relevel(fct_recode(Eng_little, 
                                      Immersion = "bileng",
                                      Immersion = "monoeng", 
                                      Immersion = "lot",
                                      `Non-Immersion` = "little", 
                                      Other = "other"),
                                      "Immersion","Other")) %>%
  group_by(age, start_binned, language_status, education) %>%
  summarise(correct = mean(correct), 
            theta= mean(fscore),
            n = n()) %>%
  filter(n > 10)

ggplot(ms,  
       aes(x = age, y = correct, col = start_binned)) +
  # geom_line() + 
  geom_point(aes(size = log10(n)), alpha = .2) + 
  facet_grid(education~language_status) + 
  geom_smooth(aes(group = start_binned), se=FALSE, span = 1) +
  theme_few() +
  scale_color_discrete(name = "Initial Exposure Age") +
  xlab("Age") + 
  ylab("Mean Correct") +
  theme(legend.position = "bottom")
```

Zoom in on monolinguals.

```{r}
ms <- d %>%
  group_by(age, Eng_little, education) %>%
  summarise(correct = mean(correct), 
            theta= mean(fscore),
            n = n()) %>%
  filter(n > 10)

ggplot(filter(ms, Eng_little == "monoeng", !is.na(education)),  
       aes(x = age, y = theta, col = education)) +
  # geom_line() + 
  geom_point(aes(size = log10(n)), alpha = .2) + 
  # facet_grid(education~language_status) + 
  geom_smooth(aes(group = education), se=FALSE, span = .75) +
  theme_few() +
  # scale_color_discrete(name = "Initial Exposure Age") +
  xlab("Age at Test") + 
  ylab("Latent Ability") +
  scale_color_ptol(name = "Education") + 
  theme(legend.position = "bottom", ) + 
  guides(color=guide_legend(ncol=2)) + 
  ylim(-.25, .5)

```

# Plot for TiCS spotlight

```{r}
a <- ggplot(d_hist, 
       aes(x = fscore, fill = language_status)) + 
  geom_histogram(binwidth = .25) + 
  theme_few() + 
  scale_fill_ptol(name = "") + 
  scale_y_continuous(name = "Number of Participants", 
                     breaks = c(1e5,2e5,3e5,4e5)) +
  xlab("Latent Ability (IRT Model Parameter)") + 
  xlim(-3.5, .8) + 
  theme(legend.position = "bottom") + 
  guides(fill = guide_legend(ncol = 2))

b <- ggplot(filter(ms, Eng_little == "monoeng", 
                   !is.na(education), education != "< High School"),  
       aes(x = age, y = theta, col = education)) +
  # geom_line() + 
  geom_point(aes(size = log10(n)), alpha = .2) + 
  geom_smooth(aes(group = education), se = FALSE, span = .6) +
  theme_few() +
  xlab("Age at Test (Years)") + 
  ylab("Latent Ability") +
  scale_size_continuous(guide = FALSE) + 
  scale_color_viridis(name = "", discrete = TRUE) + 
  theme(legend.position = "bottom") + 
  guides(color = guide_legend(ncol = 2)) + 
  ylim(-.25, .5)

# pdf("HTP.pdf", width = 8, height = 4)
cowplot::plot_grid(a,b, labels = c("A","B"))
# dev.off()
```

